---
title: "Eicosanoids mediators of hypertension"
author: "Joonatan Palmu"
date: "`r format(Sys.time(), '%d.%m.%Y')`"
---

```{r options, include=FALSE}
knitr::opts_chunk$set(include = TRUE, echo = TRUE, message = FALSE,
                      results='asis', cache=FALSE, warning=FALSE)
options(knitr.kable.NA = "")
now <- format(Sys.time(), '%Y%m%d-%H%M%S')
```

```{r read last session, include = FALSE, eval = FALSE}
message("Reading file ", paste0("rds/", sort(list.files("rds"), decreasing = TRUE)[1]))
load(paste0("rds/", sort(list.files("rds"), decreasing = TRUE)[1]))
```

```{r Remove temporary data, include = FALSE}
unlink("cache", recursive=TRUE)
dir.create("cache", showWarnings = FALSE)
dir.create("report", showWarnings = FALSE)
```

# Libraries

<details>
  <summary>Open/Close</summary>

```{r libraries, results = 'hide'}
library(knitr)
library(tidyr)
library(broom)
library(dplyr)
library(parallel)
library(xtable)
library(glmnet)
library(ggplot2)
library(scales)
library(magrittr)
library(forestplot)
library(survival)
library(png)
library(data.table)
library(tableone)
library(flextable)
library(ggrepel)
library(corrplot)
library(purrr)
library(gplots)
library(ggcorrplot)
library(Rtsne)
library(RColorBrewer)
library(gridExtra)
library(dendextend)
library(ggpubr)
library(readr)
```


</details>

# Functions

<details><summary>Functions</summary>

```{r Functions}
sourcefiles <- c("articleone-importer.R",
                 "articleone-functions.R",
                 "articleone-replication.R",
                 "articleone-riskscore.R",
                 "articleone-plots.R",
                 "articleone-tables.R")
```

```{r import files, echo = FALSE}
for (f in sourcefiles) {
    source(f)
}
```

```{r embbed files, echo = FALSE}
xfun::embed_files(c("rrnmr.Rmd", sourcefiles))
```

</details>

# Data import

```{r import data}
dset <- importdata(replacenawithmin = TRUE, normalize = TRUE)
```

```{r remove, eval = FALSE}
abund <- readRDS("eicdata/metabolites/fr02-eicosanoids-MAD.rds")
meta <- readRDS("eicdata/metabolites/fr02-biodatacoreImport-meta.rds")

dat <- merge(meta, abund$data, by.y = "key", by.x = "sample_run_id", all = TRUE)

write.table(dat,
            file = "FINRISK_EIC_MAD_Norm_EIC_Only.tsv",
            quote=FALSE,
            sep='\t',
            col.names = TRUE,
            row.names = FALSE)

```


# Characteristics

```{r Characteristics}
characteristics(dset$data.full) %>% kable
```

# Linear models

## FR02

```{r Linear models}
linear.lmrank <- calculateglm(dset = dset$data.full, 
                                           loops = dset$metabolite.mzids,
                                           responses = c2l("MAP", "SBP", "DBP", "PP", "HT"),
                                           covariates = c("AGE", "sex", "BMI", "curr_smk", "curr_diab",
                                                          "HRX", "plate")) %>%
    mutate(mzrt = pub.mzrt(term, dset$metabolite.names),
           qval = bonf.adjust(p.value, n=5))

saveRDS(linear.lmrank, file = "linear.lmrank.Rds")
```

Total number of `r dset$metabolite.mzids %>% length` eicosanoids
available and number of eicosanoids with significant associations is
`r pull.signf(linear.lmrank) %>% length`.

Number of associations per BP indice

```{r print number of associations per BP indice}
lapply(c2l("MAP", "SBP", "DBP", "PP", "HT"), function(x)
    linear.lmrank %>% filter(response == x)) %>%
    purrr::map_df(., ~as.data.frame(.x)) %>%
    group_by(response) %>%
    summarize(n = n(),
              significant = sum(qval < 0.05),
              proportion = significant/n,
              positive = sum(qval < 0.05 & estimate > 0),
              negative = sum(qval < 0.05 & estimate < 0)) %>%
    kable
```


Effect size of significant associations in FR02

```{r print linear results}
linear.lmrank %>%
    filter(qval < 0.05) %>%
    select(term, response, estimate) %>%
    mutate(estimate = round(estimate, 3)) %>%
    spread(response, estimate) %>% 
    kable
```

Number of significant associations to map

```{r number of associations for map}
filter.model(linear.lmrank, plimit = 0.05, filterterm = "SBP") %>%
    dplyr::mutate(sign = ifelse(estimate >0, 1, -1)) %>%
    group_by(sign) %>%
    summarize(n = n()) %>%
    mutate(prop = round(n/sum(n)*100, 1)) %>%
    kable
```


## Manhattan plot

For FR02 bonferroni limit for significance is `r 0.05/length(dset$metabolite.mzids)`.

```{r manhattan descriptions}
manhattanplot <- plot.manhattanplot(filter.model(linear.lmrank),
                                    bonfp = 0.05/length(dset$metabolite.mzids),
                                    nlabels = 0)
ggsave(file = "cache/manhattanplot.png", plot = manhattanplot, height = 6, width = 7, dpi = 300)
```

<img src = "cache/manhattanplot.png" />

## Heatmap of metabolites

Correlation matrix

```{r correlation matring using ggplot2}
eicosanoidcorrelation <- spearmancorrelation(dset,
                                          pull.signf(linear.lmrank, filterterm = "SBP"))

p.mat <- cor_pmat(eicosanoidcorrelation)

ggcorrelation <- ggcorrplot(eicosanoidcorrelation,
                            p.mat = p.mat,
                            hc.order = TRUE,
                            type = "full",
                            insig = "blank",
                            pch = 4,
                            pch.col = "gray",
                            show.legend = TRUE) +
    scale_fill_gradientn(name = "Spearman\ncorrelation",
                         colors = c("blue", "white", "red"),
                         breaks = c(-1, -0.5, 0, 0.5, 1),
                         limits = c(-1.1, 1.1)) +
    theme(panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),
          panel.background = element_blank(),
          axis.line = element_line(colour = "black")) +
    theme(legend.title = element_text(size = 30),
          legend.text = element_text(size = 24),
          legend.key.width = unit(14, "mm"),
          legend.key.height = unit(20, "mm"),
          axis.text.x=element_blank(),
          axis.text.y=element_blank())

ggsave(file = "cache/heatmap.png", plot = ggcorrelation, height = 14, width = 18, dpi = 300)
```
<img src = "cache/heatmap.png" />

# Multivariate models

Calculating three regularisation models

```{r Stepwise calculation, results = 'hide'}
regularisation <- regularisationmodel(dset$data.full,
                                      pull.signf(linear.lmrank),
                                      bonfp = 0.05/length(dset$metabolite.mzids))
```

Number of eicosanoids in three regularisation models

```{r number of eicosanoids in regularisation}
purrr::map_df(regularisation, ~as.data.frame(.x), .id="id") %>%
    dplyr::filter(grepl("mzid", term)) %>%
    group_by(id) %>%
    summarize(n = n()) %>%
    spread(id, n) %>%
    kable
```

Regularisation results

```{r regularisation results}
purrr::map_df(regularisation, ~as.data.frame(.x), .id="id") %>%
    dplyr::filter(grepl("mzid", term)) %>%
    select(term, id, estimate) %>%
    spread(id, estimate) %>%
    mutate_if(is.numeric, round, 3) %>%
    kable
```

Results for Bonferroni forward selection

```{r regression bonf kable}
regularisation$fwdbonf %>%
    filter(grepl("mzid_", term)) %>%
    select(term, estimate, conf.low, conf.high, p.value) %>%
    mutate_if(is.numeric, round, 3) %>%
    kable
```

# Replication

```{r results for the six eicosanoids}
compare.eicosanoids <- mygrep(regularisation$fwdbonf$term, word = "mzid")
forest.fr02 <- compare.fr02(linear.lmrank, compare.eicosanoids)
forest.fhs <- compare.fhs()
```

```{r Comparing six-eicosanoid results}
full_join(forest.fr02, forest.fhs, by = "term", suffix = c(".fr02", ".fhs")) %>%
    rename(mzrt = mzrt.fr02, name = name.fr02) %>%
    select(mzrt, name, mean_ci.fr02, mean_ci.fhs) %>% kable
```

```{r Plot the forest plot image}
replicationforestplot(forest.fr02, forest.fhs, file = "cache/forestmap-comparison.png")
```

<img src = "cache/forestmap-comparison.png" />

# Eicosanoid profiles

## FINRISK

```{r Calculate risk scores}
rset <- getriskset(dset$data.full, regularisation)
```

```{r Risk models}
riskmodel.fr02 <- lapply(list("unadjusted" = c(),
                              "adjusted" = c("AGE", "sex", "BMI", "curr_smk", "curr_diab")), function(adj)
                                  calculateglm(dset = rset,
                                               loops = c("fwdbonf_riskpersd", "fwdbonf_riskclass"),
                                               covariates = adj,
                                               responses = c("SBP", "HT"),
                                               filter = "risk")) %>%
    purrr::map_df(~as.data.frame(.x), .id="adjustment")
riskmodel.fhs <- riskmodel.fhs()
```

P for trend

```{r risk models p for trend}
calculateglm(dset = rset %>% mutate(fwdbonf_riskclass = as.numeric(fwdbonf_riskclass)),
             loops = "fwdbonf_riskclass",
             covariates = c("AGE", "sex", "BMI", "curr_smk", "curr_diab"),
             responses = c("SBP", "HT"),
             filter = "risk") %>%
    mutate(p.value = pub.p(p.value)) %>%
    select(response, term, p.value) %>%
    kable
```

Risk model results for fr02

```{r risk model results}
riskmodel.fr02 %>%
    select(adjustment, term, response, estimate, conf.low, conf.high, p.value) %>%
    kable
```

```{r risk model hypertensive per class}
rset %>% group_by(fwdbonf_riskclass) %>% summarize(htn = sum(HT == 1), n=n()) %>% kable
```

## FHS

```{r replication for FHS}
riskmodel.fhs %>%
    select(adjustment, term, response, estimate, conf.low, conf.high, p.value) %>%
    kable
```

## Comparing results

```{r Plot risk models}
plot.riskmodel(riskmodel.fhs,
               riskmodel.fr02,
               variable = "HT",
               ylab = "Odds for hypertension (95% CI)",
               file = "cache/riskforest-htn.png")
```

<img src = "cache/riskforest-htn.png" />

```{r Plot risk models SBP}
plot.riskmodel(riskmodel.fhs,
               riskmodel.fr02,
               variable = "SBP",
               ylab = "Effect size for systolic BP (95% CI)",
               file = "cache/riskforest-sbp.png")
```

<img src = "cache/riskforest-sbp.png" />

```{r proportions, eval = FALSE, include = FALSE}
risksetpre.fr02  <- rset %>%
    group_by(fwdbonf_riskclass) %>%
    summarize(ht = sum(HT == 1), nt = sum(HT == 0), n = ht+nt) %>%
    mutate(cohort = "FINRISK")

risksetpre.fhs <- rset %>%
    sample_n(2000) %>%
    group_by(fwdbonf_riskclass) %>%
    summarize(ht = sum(HT == 1), nt = sum(HT == 0), n = ht+nt) %>%
    mutate(cohort = "FHS")

risksetpre <- rbind(risksetpre.fr02, risksetpre.fhs) %>%
    rename(class = fwdbonf_riskclass) %>%
    gather(key, value, ht, nt, -cohort, -n) %>%
    mutate(key = relevel(as.factor(key), ref = "nt"))

g.proportion <- ggplot(risksetpre,
                       aes(x = class,
                           y = value/n,
                           fill = key)) +
    facet_wrap(~cohort) +
    geom_bar(stat = 'identity') +
    scale_x_discrete(labels=c("1" = "Q1",
                              "2" = "Q2",
                              "3" = "Q3",
                              "4" = "Q4")) +
    scale_fill_manual(name = "",
                      labels = c("Normotensive", "Hypertensive"),
                      breaks = c("nt", "ht"),
                      values = c("blue", "red")) +
    scale_y_continuous(limits=c(0, 1), breaks = c(0, 0.5, 1)) +
    xlab("") +
    ylab("") +
    theme_classic() +
    theme(strip.background = element_blank(),
          strip.placement = "outside")

ggsave(file = "cache/riskforest-prop.png", plot = g.proportion, height = 3, width = 4, dpi = 300)
```

# Kaplan-Mayer plot

```{r kaplanmayer}
library(emmeans)
library(survival)
library(survminer)

res.cox <- coxph(Surv(dthtime, dth) ~ fwdbonf_riskclass + AGE + sex +
                     tot_chol + hdl_chol + curr_smk + curr_diab +
                     HRX + MAP,
                 data = rset)

coxret %>% tidy(exponentiate = TRUE)

fit2 <- survfit(Surv(dthtime, dth) ~ fwdbonf_riskclass, data = rset)
ggsurvplot(fit2, data = rset, pval = TRUE, ylim = c(0.8, 1))

fit2 <- survfit(Surv(CVD_AGEDIFF, CVD) ~ fwdbonf_riskclass, data = rset %>% filter(PREVAL_CVD == 0))
ggsurvplot(fit2, data = rset, pval = TRUE, ylim = c(0.8, 1))

colnames(rset) %>% mygrep(word = "cvd")
```



# Supplement

## Linear and logistic regression results in FR02

```{r linear results for supplement}
linear.lmrank.names <-
    dset$metabolite.names %>%
    mutate(name = gsub("Novel EIC_", "Novel-", name),
           name = gsub("EIC_", "Putative-", name),
           name = gsub("Eicosanoid_", "", name),
           name = gsub("FFA", "", name),
           name = gsub(" *\\[M-H\\]", "", name),
           name = gsub("[ _]", "", name),
           name = gsub(" ", "-", name),
           name = gsub(";", "; ", name),
           name = gsub("alpha", "α", name))
    
sup.linres <- pub.lmrank(linear.lmrank) %>%
    right_join(linear.lmrank.names, ., by = c("mzid" = "term")) %>%
    select(-mzid)

typology.tbls2 <- data.frame(
  col_keys = colnames(sup.linres),
  what = c( "", "", "",
           rep("MAP", 2), 
            rep("Systolic BP", 2), 
            rep("Diastolic BP", 2), 
            rep("Pulse pressure", 2), 
            rep("Hypertension", 2)),
  measure = c("Eicosanoid", "M/Z", "RT", rep(c("β±SE", "p"), 5)),
  stringsAsFactors = FALSE)

typologyformatter(data = sup.linres, font = 9, typology = typology.tbls2)
```

```{r Write docx to file, results = FALSE, include = FALSE}
save.image(file = paste0("rds/session-", now, ".Rdata"))
```

